---
title: "Predictability of helminth parasite host range using information on geography, host traits and parasite community structure"
author: ""
output:
  pdf_document: default
  html_document:
    fig_height: 6
    fig_width: 6
    highlight: tango
    theme: journal

---


The following code was used to create predictive boosted regression models of parasite species distributions among hosts. See original citation for details of the analysis:

> Dallas, T, AW Park, JM Drake. 2016. Predictability of helminth parasite host range using information on geography, host traits and parasite community structure. _Parasitology_ doi:10.1017/S0031182016001608



The data provided are from FishPest (Strona and Lafferty 2012) and FishBase (Froese and Pauly 2000).

> Froese, R. and D. Pauly, Editors. 2000. FishBase 2000: concepts, design and data sources. ICLARM, Los Baños, Laguna, Philippines. 344 p.

> Strona, G and KD Lafferty. "FishPEST: an innovative software suite for fish parasitologists." Trends in parasitology 28.4 (2012): 123.


### Load required packages

```{r echo=FALSE, message=FALSE, comment=FALSE}
library(gbm);
library(ROCR);
library(randomForest);
library(vegan);
library(plyr);
library(dplyr)
```



### Load data from FishPest

```{r}
fishData <- read.csv('fishPest.csv', stringsAsFactors = FALSE)
```


```{r eval=F, echo=TRUE}
#remove first 3 variables relating to parasite taxonomy
pest <- fishData[,-c(1:3)]
pest[which(pest=='na', arr.ind=TRUE)] <- NA
pest[, 7:ncol(pest)] <- apply(pest[,7:ncol(pest)], 2,  as.numeric)
#find parasites sampled 15 or more times
parasites <- names(which(table(pest$P_SP) > 10))
```



## Calculate host taxonomic eigenvectors

```{r eval=F, echo=T}

fishData$H_G <- sapply(as.character(fishData$H_SP), function(x){unlist(strsplit(x, ' '))[1]})

uniqueFishData <- unique(fishData[,c('H_C','H_O','H_F','H_G')])

hostTaxDist <- taxa2dist(uniqueFishData, varstep=TRUE)
hostTaxDist <- as.data.frame(as.matrix(hostTaxDist))
avgTaxDist <- rowMeans(hostTaxDist)

hostEig <- ape::pcoa(hostTaxDist)

uniqueFishData$pc1 <- hostEig$vectors[,1]
uniqueFishData$pc2 <- hostEig$vectors[,2]
uniqueFishData$pc3 <- hostEig$vectors[,3]
uniqueFishData$pc4 <- hostEig$vectors[,4]
uniqueFishData$pc5 <- hostEig$vectors[,5]
```


```{r eval=FALSE, echo=FALSE}
write.csv(uniqueFishData, file='uniqueFishData.csv', row.names=FALSE)
```


#### Training models on host trait data

```{r, eval=FALSE, echo=T, message=FALSE}
# storage for model outputs
 baseline.auc <- aso.auc <-  brt.perfAUC <- brt.perfAUC.geog <- brt.perfAUC.trait <- brt.perfAUC.parcom <- brt.perfAUC.hostTax <-vector()

 brtModel <- brt.best.iter <- brt.preds <- brt.perf <- list()
 brtModel.geog <- brt.best.iter.geog <- brt.preds.geog <- brt.perf.geog <- list()
 brtModel.trait <- brt.best.iter.trait <- brt.preds.trait <- brt.perf.trait <- list()
 brtModel.parcom <- brt.best.iter.parcom <- brt.preds.parcom <- brt.perf.parcom <- list()
 brtModel.hostTax <- brt.best.iter.hostTax <- brt.preds.hostTax <- brt.perf.hostTax <- list()

# form the data into an interaction matrix,
# and determine parasite species richness
 intMat <- table(pest$H_SP, pest$P_SP)
 psr <- rowSums(intMat)

 hostData <- unique(pest[,c(2:ncol(pest))])
 hostData2 <- hostData[-which(is.na(hostData$GEO)), ]
#encode biogeographic region as dummy variable
 hostData2$AFR <- as.numeric(hostData2$GEO == 'AFR')
 hostData2$PAL <- as.numeric(hostData2$GEO == 'PAL')
 hostData2$ANT <- as.numeric(hostData2$GEO == 'ANT')
 hostData2$NEO <- as.numeric(hostData2$GEO == 'NEO')
 hostData2$AUS <- as.numeric(hostData2$GEO == 'AUS')
 hostData2$IND <- as.numeric(hostData2$GEO == 'IND')
 hostData2$NEA <- as.numeric(hostData2$GEO == 'NEA')
#remove GEO from hostData2
 hostData2$GEO <- NULL
 hostData2$H_G <- sapply(as.character(hostData2$H_SP), function(x){unlist(strsplit(x, ' '))[1]})
 hostData2 <- full_join(hostData2, uniqueFishData, by = c('H_C','H_O','H_F','H_G'))
 hostData2$H_G <- NULL
 hostData2$H_C <- NULL
 hostData2$H_O <- NULL
 hostData2$H_F <- NULL

#condense hostData2 down to 1 unique entry per host species
 hosts <- unique(hostData2$H_SP)
 hostData3 <-hostData2[1,]
 colnames(hostData3) <- colnames(hostData2)

 for(g in 1:length(hosts)){
   temp <- hostData2[which(hostData2[,'H_SP'] == hosts[g]),]
   if(nrow(temp) == 1){
     hostData3 <- bind_rows(hostData3, temp)
   }else{
     temp2 <- temp[1,]
     temp2[13:19] <- colSums(temp[,13:19])
     hostData3 <- bind_rows(hostData3, temp2)
   }
   print(g)
 }


# tiny tweaks to data format
 hostData4 <- as.data.frame(hostData3[-1,])
 hostData4$F <- NULL;
 hostData4$B <- NULL;
 hostData4$M <- NULL;
 hostData4$H_SP <- unique(hostData2$H_SP)
```




```{r}
#' This function calculates variable importance as I define it in the supplement.
#'
#'
#' @param gbmModel: a trained model object from `gbm`
#' @param newdata: a set of data with identical column structure to the training data set used in `gbmModel`
#' @param prezTE: presence values for the test data set (used to assess model accuracy)
#' @param AUC: overall model AUC, used to standardize change in model accuracy
#' @param numSims: number of times to permute each predictor variable
#' @param dummyCols: which columns correpsond to the same variable (used to handle biogeographic region)
#' @param best.iter: what was the optimal number of trees in the `gbmModel`?


getPerformance <- function(gbmModel, newdata, prezTE, AUC,
  numSims=1000, dummyCols = NULL, best.iter){
  require(ROCR)
  if(is.null(dummyCols)){cols <- 1:ncol(newdata)}
  if(!is.null(dummyCols)){cols <- (1:ncol(newdata))[-c(dummyCols)]}
  out <- vector()
  for(i in 1:length(cols)){
      newdata2 <- newdata
      relRed <- vector()
      colcol <- cols[i]
      for(n in 1:numSims){
         newdata2[,i] <- newdata2[sample(1:nrow(newdata2)), colcol]
         relRed[n] <- unlist(performance(prediction(predict(gbmModel, newdata = newdata2, n.trees = best.iter), prezTE),'auc')@y.values) / AUC
      }
      out[i] <- mean(relRed)
    }
   names(out) <- colnames(newdata2)[cols]
    if(!is.null(dummyCols)){
      newdata2 <- newdata
      relRed <- vector()
      for(n in 1:numSims){
        newdata2[,dummyCols] <- newdata2[sample(1:nrow(newdata2)), dummyCols]
        relRed[n] <- unlist(performance(prediction(predict(gbmModel, newdata = newdata2, n.trees = unlist(best.iter)), prezTE),'auc')@y.values) / AUC
      }
     out[i+1] <- mean(relRed)
    }
  return(out)
}
```





### Model training

```{r eval=FALSE, echo=TRUE}
parasiteVar <- vector()

for(i in 1:length(parasites)){
 #create presence vector
  presence <- rep(0, length(hostData4$H_SP))
  names(presence) <- hostData4$H_SP
  presence[which(names(presence) %in%
    pest[which(pest$P_SP == parasites[i]), 'H_SP'])] <- 1
 #calculates PCA on parasite communities after removing focal parasite species
  intMat.trunc <- intMat[,which(colSums(intMat) > 10)]
  pca.parasite <- princomp(intMat.trunc[,-which(colnames(intMat.trunc) ==
                                                  parasites[i])])

  parasiteVar[i] <- cumsum(pca.parasite$sdev^2 / sum(pca.parasite$sdev^2))[5]
  pca.parasite <- pca.parasite$scores[,1:5]
  # assign PCA loadings to hosts (a bit clunky)
  hostPCA <- matrix(0, ncol=5, nrow=length(presence))
  hostPSR <- rep(0, length(presence))
  names(hostPSR) <- names(presence)
  rownames(hostPCA) <- names(presence)
  hostNames <- rownames(hostPCA)
  # loop through hosts and attribute PCA loadings and parasite
  #   species richness values to each host
  for(m in 1: length(hostNames)){
    hostPCA[which(rownames(hostPCA) == hostNames[m]),] <-
             pca.parasite[which(rownames(pca.parasite) == hostNames[m]),]
    hostPSR[which(names(hostPSR) == hostNames[m])] <- psr[which(names(psr)==hostNames[m])]
  }
  hostPCA <- cbind(hostPCA, hostPSR)
  #Only train on some of the absences, since they're not true absences
  cutDown <- c(which(presence == 1), sample(which(presence == 0),
                                            20 * sum(presence)))
  presence1 <- presence[cutDown]
  hostPCA1 <- hostPCA[cutDown, ]
  hostData5 <- cbind(hostData4[cutDown,], hostPCA1)
  if(any(is.na(hostData5$H_SP))){
      ind <- which(is.na(hostData5$H_SP))
      hostData5 <- hostData5[-ind,]
      presence1 <- presence1[-ind]
  }
  #Impute the data
  impDat <- rfImpute(hostData5[,-1], presence1)

  # This makes sure that the test set contains at
  # least 4 hosts on which the parasite actually occurs
  inds <- sample(1:nrow(impDat), 0.7 * nrow(impDat))
  if(sum(presence1[inds]) < 3){
    inds[1:4] <- which(presence1 == 1)[1:4]
  }
  if(sum(presence1[-inds])==0){
    inds <- inds[-which(inds==1)[1]]
  }
  #Set up a prelim train set and a test set
  train <- impDat[inds,-1]
  test  <- impDat[-inds,-1]
 #Presences for training (prezTR) and test (prezTE) sets
  prezTR <- presence1[inds]
  prezTE <- presence1[-inds]
 #Baseline expectations and null models
  baseline.auc[i] <- unlist(performance(prediction(sample(presence[-inds], length(presence[-inds])), presence[-inds]), 'auc')@y.values)
  # A single occurrence prediction
  aso.auc[i] <- unlist(performance(prediction(c(1,rep(0, length(prezTE)-1)),
                                               prezTE), 'auc')@y.values)
#   ----
  # Models trained on geographic variables
  geog <- c(6:15)
  # error handling, just in case the gbm model can't be fit
  temp <- tryCatch(
     gbm(prezTR ~ ., data = train[,geog], n.trees = 100,
         interaction.depth = 4, distribution = 'bernoulli',
         n.cores=5), error = function(e){e}
  )

  if(!inherits(temp, 'error')){
   brt.best.iter.geog[[i]] <- gbm.perf(temp, method="OOB")
   brt.preds.geog[[i]] <- prediction(predict(temp, newdata=test[,geog],
                                          n.trees=brt.best.iter.geog[[i]]), prezTE)
   brt.perf.geog[[i]]  <- performance(brt.preds.geog[[i]],"tpr","fpr")
   brt.perfAUC.geog[i] <- unlist(performance(brt.preds.geog[[i]], 'auc')@y.values)
   brtModel.geog[[i]] <- summary(temp, order=FALSE)

#   ----
 # Models trained on host traits
  trait <- c(1:5)
  temp <- gbm(prezTR ~ ., data = train[,trait], n.trees = 50000,
              interaction.depth = 4, distribution = 'bernoulli', n.cores=5)
  brt.best.iter.trait[[i]] <- gbm.perf(temp, method="OOB")
  brt.preds.trait[[i]] <- prediction(predict(temp, newdata=test[,trait],
                                         n.trees=brt.best.iter.trait[[i]]), prezTE)
  brt.perf.trait[[i]]  <- performance(brt.preds.trait[[i]],"tpr","fpr")
  brt.perfAUC.trait[i] <- unlist(performance(brt.preds.trait[[i]], 'auc')@y.values)
  brtModel.trait[[i]] <- summary(temp, order=FALSE)


#   ----
  # Models trained on parasite community variables
  parcom <- c(21:26)
  temp <- gbm(prezTR ~ ., data = train[,parcom], n.trees = 50000,
              interaction.depth = 4, distribution = 'bernoulli', n.cores=5)

  brt.best.iter.parcom[[i]] <- gbm.perf(temp, method="OOB")
  brt.preds.parcom[[i]] <- prediction(predict(temp, newdata=test[,parcom],
                                         n.trees=brt.best.iter.parcom[[i]]), prezTE)
  brt.perf.parcom[[i]] <- performance(brt.preds.parcom[[i]],"tpr","fpr")
  brt.perfAUC.parcom[i] <- unlist(performance(brt.preds.parcom[[i]], 'auc')@y.values)
  brtModel.parcom[[i]] <- summary(temp, order=FALSE)

#   ----
  # Models trained on host taxonomic variables
  hostTax <- c(16:20)
  temp <- gbm(prezTR ~ ., data = train[,hostTax], n.trees = 50000,
              interaction.depth = 4, distribution = 'bernoulli', n.cores=5)
  brt.best.iter.hostTax[[i]] <- gbm.perf(temp, method="OOB")
  brt.preds.hostTax[[i]] <- prediction(predict(temp, newdata=test[,hostTax],
                                       n.trees=brt.best.iter.hostTax[[i]]), prezTE)
  brt.perf.hostTax[[i]] <- performance(brt.preds.hostTax[[i]],"tpr","fpr")
  brt.perfAUC.hostTax[i] <- unlist(performance(brt.preds.hostTax[[i]], 'auc')@y.values)
  brtModel.hostTax[[i]] <- summary(temp, order=FALSE)

#   ----
   # Models trained on all variables
  temp <- gbm(prezTR ~ ., data = train, n.trees = 50000,
                        interaction.depth = 4, distribution = 'bernoulli', n.cores=5)
  brt.best.iter[[i]] <- gbm.perf(temp, method="OOB")
  brt.preds[[i]] <- prediction(predict(temp, newdata=test,
                                       n.trees=brt.best.iter[[i]]), prezTE)
  brt.perf[[i]] <- performance(brt.preds[[i]],"tpr","fpr")
  brt.perfAUC[i] <- unlist(performance(brt.preds[[i]], 'auc')@y.values)
  brtModel[[i]] <- summary(temp, order=FALSE)
 }
 print(i)
}


 # model accuracy
 acc <- data.frame(base = unlist(baseline.auc),
    oneShot = aso.auc,
    brt.geog = brt.perfAUC.geog,
    brt.trait = brt.perfAUC.trait,
    brt.parcom = brt.perfAUC.parcom,
    brt.hostTax = brt.perfAUC.hostTax,
    brt.all = brt.perfAUC)
```







### Model result formatting and such

```{r, eval=FALSE, echo=FALSE}
#Extracting the BRT predictor variable importance values
brtResults.traits <- matrix(0, ncol=length(brtModel), nrow = 5)
brtResults.geo <- matrix(0, ncol=length(brtModel), nrow = 10)
brtResults.pars <- matrix(0, ncol=length(brtModel), nrow = 6)
brtResults.hostTax <- matrix(0, ncol=length(brtModel), nrow=5)
brtResults.all <- matrix(0, ncol=length(brtModel), nrow = 26)

for(i in 1:length(brtModel)){

 if(is.null(brtModel[[i]])){

 }else{
   brtResults.traits[,i] <- brtModel.trait[[i]][order(brtModel.trait[[i]][,1]),2]
   brtResults.geo[,i]    <- brtModel.geog[[i]][order(brtModel.geog[[i]][,1]),2]
   brtResults.hostTax[,i]    <- brtModel.hostTax[[i]][order(brtModel.hostTax[[i]][,1]),2]
   brtResults.pars[,i]   <- brtModel.parcom[[i]][order(brtModel.parcom[[i]][,1]),2]
   brtResults.all[,i]    <- brtModel[[i]][order(brtModel[[i]][,1]),2]
   print(i)
 }
}

rownames(brtResults.traits)  <- sort(brtModel.trait[[1]][,1])
rownames(brtResults.geo)     <- sort(brtModel.geog[[1]][,1])
rownames(brtResults.hostTax) <- sort(brtModel.hostTax[[1]][,1])
rownames(brtResults.pars)    <- sort(brtModel.parcom[[1]][,1])
rownames(brtResults.all)     <- sort(brtModel[[1]][,1])

save(brtResults.traits, brtResults.geo, brtResults.hostTax,
    brtResults.pars, brtResults.all, acc, file='accuracyAndBars.RData')
```




```{r}
brtModel2 <- t(ldply(brtModel, function(x){return(x[,2])}))
brtModel.geog2 <- t(ldply(brtModel.geog, function(x){return(x[,2])}))
brtModel.trait.sc <- t(ldply(brtModel.trait, function(x){return(x[,2])}))
brtModel.parcom.sc <- t(ldply(brtModel.parcom, function(x){return(x[,2])}))
brtModel.hostTax2 <- t(ldply(brtModel.hostTax, function(x){return(x[,2])}))
```


```{r}
rownames(brtModel2) <- brtModel[[1]][,1]
rownames(brtModel.geog2) <- brtModel.geog[[1]][,1]
rownames(brtModel.trait.sc) <- brtModel.trait[[1]][,1]
rownames(brtModel.parcom.sc) <- brtModel.parcom[[1]][,1]
rownames(brtModel.hostTax2) <- brtModel.hostTax[[1]][,1]

brtModel.geog2 <- rbind(brtModel.geog2, biogeog = colSums(brtModel.geog2[c(4:10),], na.rm=TRUE))
brtModel.geog.sc <- brtModel.geog2[-c(4:10),]

brtModel2 <- rbind(brtModel2, biogeog = colSums(brtModel2[c(9:15),]))
brtModel.sc <- brtModel2[-c(9:15),]
```







## Figure 1

```{r}
# The `add.alpha` function was taken from a gist by mages (https://gist.github.com/mages/5339689)

add.alpha <- function(col, alpha=1){
  if(missing(col))
    stop("Please provide a vector of colours.")
  apply(sapply(col, col2rgb)/255, 2,
                     function(x)
                       rgb(x[1], x[2], x[3], alpha=alpha))
}


cols <- c('#d1006a', '#6ad100', '#006ad1', '#FF9F1C')

layout(matrix(c(1,2),ncol=2), width=c(1,0.25))

brtMean.all <- rowMeans(brtModel.sc, na.rm=TRUE)
brtSE.all <- apply(brtModel.sc, 1, sd, na.rm=TRUE) #/ sqrt(ncol(brtModel.sc))

traits <- data.frame(var = c("MaxL", "K", "Y", "Ym", "T", "AOO", "LAT", "LON", "pc1", "pc2",
                            "pc3", "pc4", "pc5", "V1", "V2", "V3", "V4", "V5", "hostPSR",
                            "biogeog"),
                    name = c("Max Length", "Host growth rate", "Host lifespan",  "Host age at maturity",
                            "Host trophic level", "Area of occupancy",  "Latitude", "Longitude", "TaxonomyPCoA.1",  "TaxonomyPCoA.2","TaxonomyPCoA.3",  "TaxonomyPCoA.4",  "TaxonomyPCoA.5",  "ParasitePCA.1", "ParasitePCA.2", "ParasitePCA.3", "ParasitePCA.4", "ParasitePCA.5", "Parasite species richness", "Biogeographic region"),
                    group = c(1,1,1,1,1,2,2,2,4,4,4,4,4,3,3,3,3,3,3,2),                                                 brtMean.all = brtMean.all, brtSE.all = brtSE.all)

traits <- traits[order(traits$brtMean.all),]
par(mar=c(4,12,1,1))
plot(traits$brtMean.all, 1:20, pch=16, xlim=c(-2,25), ylim=c(0.75,20.25), las=1,
  xlab='', ylab='', yaxt='n', col=cols[traits$group], tck=0.01)
segments(x0=traits$brtMean.all - traits$brtSE.all, x1=traits$brtMean.all + traits$brtSE.all,
  y0=1:20, col=add.alpha(cols[traits$group],0.5), lwd=3)
points(traits$brtMean.all, 1:20, pch=16, cex=1.5, col=cols[traits$group])
mtext("Relative contribution", side=1, line=2)
axis(2, at=1:20, labels=traits$name, las=1, tck=-0.01)

par(mar=c(4,0,1,4))
plot(x=1:6, colMeans(acc[,-2], na.rm=TRUE)[c(1,3,5,2,4,6)], pch=21, bg=c(grey(0.5,0.5), cols[c(1,4,2,3)], 1), col=1,
  las=1, xlab='', tck=0, ylab='AUC', xaxt='n', yaxt='n', cex=2, xlim=c(0,7), ylim=c(0.5, 1))
segments(x0=1:6, y0 = colMeans(acc[,-2], na.rm=TRUE)[c(1,3,5,2,4,6)] -
  2*apply(acc[,-2],2, function(x){sd(x, na.rm=TRUE)/sqrt(length(na.omit(x)))})[c(1,3,5,2,4,6)],
  y1 =  colMeans(acc[,-2], na.rm=TRUE)[c(1,3,5,2,4,6)] + 2*apply(acc[,-2],2, function(x){sd(x, na.rm=TRUE)/sqrt(length(na.omit(x)))})[c(1,3,5,2,4,6)], col=add.alpha(c(grey(0.5,0.5), cols[c(1,4,2,3)], 1), 0.5),lwd=4)
points(x=1:6, colMeans(acc[,-2], na.rm=TRUE)[c(1,3,5,2,4,6)], pch=21, bg=c(grey(0.5,0.5), cols[c(1,4,2,3)], 1), col=1, cex=2)
axis(4, at=seq(0.5,1,by=0.1), labels=seq(0.5,1,by=0.1), tck=0.01, las=1)
text(3.5, 1, 'AUC', font=2)

dev.copy(pdf, 'auc.pdf', width=8.5, height=5); dev.off()
```




### What variables are important in submodels?

```{r }
layout(matrix(c(1,3,2,4),ncol=2), height=c(1,1.1))
# (host traits)
brtMean.traits <- rowMeans(brtModel.trait.sc)
brtSE.traits <- apply(brtModel.trait.sc, 1, sd) #/ sqrt(ncol(brtModel.trait.sc))
inds.traits <- order(brtMean.traits)

brtMean.traits <- brtMean.traits[order(names(brtMean.traits))]
nms <- traits[which(traits$var %in% names(brtMean.traits[inds.traits])), c('var','name')]
names(brtMean.traits) <- nms[order(nms$var),'name']
inds.traits <- order(brtMean.traits)


par(mar=c(2,9,0.25,0.25))
plot(brtMean.traits[inds.traits], 1:5, pch=16, xlim=c(0,70), ylim=c(0.75,5.25), las=1, xlab='',
  ylab='', yaxt='n', col=cols[1], tck=0.01)
segments(x0=brtMean.traits[inds.traits] - brtSE.traits[inds.traits],
  x1=brtMean.traits[inds.traits] + brtSE.traits[inds.traits],
  y0=1:5, col=add.alpha(cols[1],0.5), lwd=3)
points(brtMean.traits[inds.traits], 1:5, pch=16, col=cols[1], cex=1.5)
text(54,0.85, expression(paste(bar(italic("AUC")), " = 0.72")), cex=1.25, font=2, col=cols[1])
axis(2, at=1:5, labels=names(brtMean.traits[inds.traits]), las=1)


# (geographic vars)
brtMean.geog <- rowMeans(brtModel.geog.sc, na.rm=TRUE)
brtSE.geog <- apply(brtModel.geog.sc, 1, sd, na.rm=TRUE) #/ sqrt(ncol(brtModel.geog.sc))
inds.geog <- order(brtMean.geog)

brtMean.geog <- brtMean.geog[order(names(brtMean.geog))]
nms <- traits[which(traits$var %in% names(brtMean.geog[inds.geog])), c('var','name')]
names(brtMean.geog) <- nms[order(nms$var),'name']
inds.geog <- order(brtMean.geog)

plot(brtMean.geog[inds.geog], 1:4, pch=16, xlim=c(0,70), ylim=c(0.75,4.25), las=1,
  xlab='', ylab='', yaxt='n', col=cols[2], tck=0.01)
segments(x0 = brtMean.geog[inds.geog] - brtSE.geog[inds.geog],
  x1 = brtMean.geog[inds.geog] + brtSE.geog[inds.geog],
  y0=1:4, col=add.alpha(cols[2],0.5),lwd=3)
points(brtMean.geog[inds.geog],1:4, pch=16, cex=1.5, col=cols[2])
text(54,0.85, expression(paste(bar(italic("AUC")), " = 0.85")), cex=1.25, font=2, col=cols[2])
axis(2, at=1:4, labels=names(brtMean.geog[inds.geog]), las=1)



par(mar=c(4,9,0.25,0.25))
# (host taxonomy)
brtMean.hostTax <- rowMeans(brtModel.hostTax2, na.rm=TRUE)
brtSE.hostTax <- apply(brtModel.hostTax2, 1, sd, na.rm=TRUE)
inds.hostTax <- order(brtMean.hostTax)

brtMean.hostTax <- brtMean.hostTax[order(names(brtMean.hostTax))]
nms <- traits[which(traits$var %in% names(brtMean.hostTax[inds.hostTax])), c('var','name')]
names(brtMean.hostTax) <- nms[order(nms$var),'name']
inds.hostTax <- order(brtMean.hostTax)

plot(brtMean.hostTax[inds.hostTax], 1:5, pch=16, xlim=c(0,70), ylim=c(0.75,5.25), las=1,
  xlab='', ylab='', yaxt='n', col=cols[4], tck=0.01)
segments(x0 = brtMean.hostTax[inds.hostTax] - brtSE.hostTax[inds.hostTax],
  x1 = brtMean.hostTax[inds.hostTax] + brtSE.hostTax[inds.hostTax],
  y0=1:5, col=add.alpha(cols[4],0.5),lwd=3)
points(brtMean.hostTax[inds.hostTax],1:5, pch=16, cex=1.5, col=cols[4])
text(54,0.85, expression(paste(bar(italic("AUC")), " = 0.84")), cex=1.25, font=2, col=cols[4])
axis(2, at=1:5, labels=names(brtMean.hostTax[inds.hostTax]), las=1)
mtext("Relative contribution", side=1, line=2)



# (parasite community)
brtMean.parcom <- rowMeans(brtModel.parcom.sc, na.rm=TRUE)
brtSE.parcom <- apply(brtModel.parcom.sc, 1, sd, na.rm=TRUE)# / sqrt(ncol(brtModel.parcom.sc))
inds.parcom <- order(brtMean.parcom)

brtMean.parcom <- brtMean.parcom[order(names(brtMean.parcom))]
nms <- traits[which(traits$var %in% names(brtMean.parcom[inds.parcom])), c('var','name')]
nms$name <- as.character(nms$name)
nms$name[which(nms$var=='hostPSR')] <- 'Parasite species \n richness'
names(brtMean.parcom) <- nms[order(nms$var),'name']
inds.parcom <- order(brtMean.parcom)

plot(brtMean.parcom[inds.parcom], 1:6, pch=16, xlim=c(0,70), ylim=c(0.75,6.25), las=1, xlab='',
  ylab='', yaxt='n', col=cols[3], tck=0.01)
segments(x0=brtMean.parcom[inds.parcom] - brtSE.parcom[inds.parcom],
  x1=brtMean.parcom[inds.parcom] + brtSE.parcom[inds.parcom],
  y0=1:6, col=add.alpha(cols[3], alpha=0.5),lwd=3)
points(brtMean.parcom[inds.parcom], 1:6, pch=16, cex=1.5,col=cols[3])
mtext("Relative contribution", side=1, line=2)
text(54,0.85, expression(paste(bar(italic("AUC")), " = 0.90")), cex=1.25, font=2, col=cols[3])
axis(2, at=1:6,labels=names(brtMean.parcom[inds.parcom]),las=1)

dev.copy(pdf, 'submodels.pdf', width=7, height=5.5);dev.off()
```




# Supplemental figures

### Does predictive power differ systematically among parasites?

```{r eval=T, echo=F}

#Number of hosts each parasite infects
hostNum <- table(pest$P_SP)
hostNum <- as.numeric(hostNum[which(names(hostNum) %in% parasites)])

#Beating the data up to get parasite 'type'
parPest <- parasites
parType <- unique(fishData[which(fishData[,'P_SP'] %in% parPest), 2:4], MARGIN=1)
parType[,1] <- gsub('[*]','',parType[,1])
parType <- unique(parType, MARGIN=1)
parType <- parType[order(parType[,3]),]
multNamed <- tail(names(sort(table(parType[,3]))), 4)
for(i in 1:length(multNamed)){
  parType <- parType[-which(parType[,3] %in% multNamed[i])[1],]
}

par(mar=c(4,4,0.5,0.5))
plot(log(hostNum), acc$brt.all, tck=0.01, las=1, xlab='log(# host species infected)',
  pch=16, ylab='Accuracy (AUC)', col=add.alpha(1,0.4), ylim=c(0.25,1), type='p')
points(log(hostNum), acc$brt.geog, pch=16, col=add.alpha(cols[1],0.4))
points(log(hostNum), acc$brt.trait, pch=16, col=add.alpha(cols[2],0.4))
points(log(hostNum), acc$brt.parcom, pch=16, col=add.alpha(cols[3],0.4))
points(log(hostNum), acc$brt.hostTax, pch=16, col=add.alpha(cols[4],0.4))
abline(h=0.5, col=1, lwd=2, lty=4)

if(any(is.na(acc))){
  inds <- which(is.na(acc$brt.all))
  hostNum2 <- hostNum[-inds]
  acc2 <- acc[-inds,]
}else{
  hostNum2 <- hostNum
  acc2 <- acc
}
lines(smooth.spline(log(hostNum2), acc2$brt.all, spar=0.5), lwd=4.25, col=1)
lines(smooth.spline(log(hostNum2), acc2$brt.all, spar=0.5), lwd=3, col=grey(0.5))
lines(smooth.spline(log(hostNum2), acc2$brt.geog, spar=0.5), lwd=4.25, col=1)
lines(smooth.spline(log(hostNum2), acc2$brt.geog, spar=0.5), lwd=3, col=cols[2])
lines(smooth.spline(log(hostNum2), acc2$brt.trait, spar=0.5), lwd=4.25, col=1)
lines(smooth.spline(log(hostNum2), acc2$brt.trait, spar=0.5), lwd=3, col= cols[1])
lines(smooth.spline(log(hostNum2), acc2$brt.parcom, spar=0.5), lwd=4.25, col=1)
lines(smooth.spline(log(hostNum2), acc2$brt.parcom, spar=0.5), lwd=3, col= cols[3])
lines(smooth.spline(log(hostNum2), acc2$brt.hostTax, spar=0.5), lwd=4.25, col=1)
lines(smooth.spline(log(hostNum2), acc2$brt.hostTax, spar=0.5), lwd=3, col= cols[4])

abline(h=0.5, col=grey(0.5,0.5), lwd=2, lty=2)
legend(4, 0.4, c('Full model', 'Parasite community', 'Geography', 'Host taxonomy', 'Host traits'), pch=16, col=c(grey(0.5), cols[c(3,2,4,1)]), bty='n')

dev.copy(pdf, width=5, height=5, 'hostNumAccuracy.pdf'); dev.off()
```







## Plotting model performance as a function of parasite taxa for each covariate group

```{r eval=TRUE, echo=TRUE,fig.height=7, fig.width=7}
layout(matrix(c(1,2,3,4,2,5), ncol=3, byrow=TRUE))
hostCol <- c("#9768DD", "#29A0EB", "#C55F30", "#C74A6C", "#66903C")

par(mar=c(1,4,2,1))
plot(as.factor(parType[,1]), acc$brt.trait, col=add.alpha(hostCol, 0.75), tck=0.01, pch=16, las=1, ylab='Accuracy (AUC)', xaxt='n', ylim=c(0.2,1), main="Host traits", outcol=grey(0.5,0.25), staplecol = hostCol, staplelwd=2, whisklty=1, whiskcol=hostCol, whisklwd=2)
abline(h=0.5, col=grey(0.5,0.5), lwd=2, lty=2)

par(mar=c(10,2,2,1))
plot(as.factor(parType[,1]), acc$brt.all, col=add.alpha(hostCol, 0.75), pch=16, tck=0.01, las=1,
  ylab='', xaxt='n', ylim=c(0.2,1), main="Full model",outcol=grey(0.5,0.25),
  staplecol = hostCol, staplelwd=2, whisklty=1, whiskcol=hostCol, whisklwd=2)

parGroups <- c('Acanthocephala', 'Cestoda', 'Monogenea', 'Nematoda', 'Trematoda')
axis(1, 1:5, parGroups, las=2, tck=0.01, padj=0, cex.axis=1.25)
abline(h=0.5, col=grey(0.5,0.5), lwd=2, lty=2)


par(mar=c(1,2,2,1))
plot(as.factor(parType[,1]), acc$brt.geog, col=add.alpha(hostCol, 0.75), tck=0.01, pch=16,
  las=1, ylab='Accuracy (AUC)', xaxt='n', ylim=c(0.2,1), main="Geography",
  outcol=grey(0.5,0.25), staplecol = hostCol, staplelwd=2, whisklty=1,
  whiskcol=hostCol, whisklwd=2)

par(mar=c(1,4,2,1))
plot(as.factor(parType[,1]), acc$brt.parcom, col=add.alpha(hostCol, 0.75), tck=0.01,pch=16,
  las=1, ylab='', xaxt='n', ylim=c(0.2,1), main="Parasite community",outcol=grey(0.5,0.25),
  staplecol = hostCol, staplelwd=2, whisklty=1, whiskcol=hostCol, whisklwd=2)
abline(h=0.5, col=grey(0.5,0.5), lwd=2, lty=2)

par(mar=c(1,2,2,1))
plot(as.factor(parType[,1]), acc$brt.hostTax, col=add.alpha(hostCol, 0.75), tck=0.01,pch=16,
  las=1, ylab='', xaxt='n', ylim=c(0.2,1), main="Host taxonomy",outcol=grey(0.5,0.25),
  staplecol = hostCol, staplelwd=2, whisklty=1, whiskcol=hostCol, whisklwd=2)
abline(h=0.5, col=grey(0.5,0.5), lwd=2, lty=2)

dev.copy(pdf, width=7, height=7, 'parAccuracy.pdf'); dev.off()
```







## How does the number of background points influence model accuracy?

```{r eval=FALSE, echo=TRUE}

getBackground <- function(parasites, bckgrnd = 10){
  parasiteVar <- vector()
  baseline.auc <- brt.auc <- vector()
  brtModel <- list()
  for(i in 1:length(parasites)){
   #create presence vector
    presence <- rep(0, length(hostData4$H_SP))
    names(presence) <- hostData4$H_SP
    presence[which(names(presence) %in% pest[which(pest$P_SP == parasites[i]), 'H_SP'])] <- 1
   #calculates PCA on parasite communities after removing focal parasite species
    intMat.trunc <- intMat[,-which(colSums(intMat) < 10)]
    pca.parasite <- princomp(intMat.trunc[,-which(colnames(intMat.trunc) ==
                                                  parasites[i])])

    parasiteVar[i] <- cumsum(pca.parasite$sdev^2 / sum(pca.parasite$sdev^2))[5]
    pca.parasite <- pca.parasite$scores[,1:5]
    # assign PCA loadings to hosts (a bit clunky)
    hostPCA <- matrix(0, ncol=5, nrow=length(presence))
    hostPSR <- rep(0, length(presence))
    names(hostPSR) <- names(presence)
    rownames(hostPCA) <- names(presence)
    hostNames <- rownames(hostPCA)
   # loop through hosts and attribute PCA loadings and parasite
   #   species richness values to each host
    for(m in 1: length(hostNames)){
      hostPCA[which(rownames(hostPCA) == hostNames[m]),] <-
             pca.parasite[which(rownames(pca.parasite) == hostNames[m]),]
      hostPSR[which(names(hostPSR) == hostNames[m])] <- psr[which(names(psr)==hostNames[m])]
    }
    hostPCA <- cbind(hostPCA, hostPSR)

    #Only train on some of the absences, since they're not true absences
    cutDown <- c(which(presence == 1), sample(which(presence == 0),
                                              min(c(bckgrnd * sum(presence), sum(presence==0)))))

    presence1 <- presence[cutDown]
    hostPCA1 <- hostPCA[cutDown, ]

    hostData5 <- cbind(hostData4[cutDown,], hostPCA1)
    if(any(is.na(hostData5$H_SP))){
        ind <- which(is.na(hostData5$H_SP))
        hostData5 <- hostData5[-ind,]
        presence1 <- presence1[-ind]
    }
    #Impute the data
    impDat <- rfImpute(hostData5[,-1], presence1)

    # This makes sure that the test set contains at
    # least 4 hosts on which the parasite actually occurs
    inds <- sample(1:nrow(impDat), 0.7 * nrow(impDat))
    if(sum(presence1[inds]) < 4){
      inds[1:4] <- which(presence1 == 1)[1:4]
    }

    #Set up a prelim train set and a test set
    train <- impDat[inds,-1]
    test  <- impDat[-inds,-1]

   #Presences for training (prezTR) and test (prezTE) sets
    prezTR <- presence1[inds]
    prezTE <- presence1[-inds]

   #Baseline expectations and null models
    baseline.auc[i] <- unlist(performance(prediction(sample(presence[-inds], length(presence[-inds])), presence[-inds]), 'auc')@y.values)

#   ----
    temp <- tryCatch(
        gbm(prezTR ~ ., data = train,  n.trees = 100,
                    interaction.depth = 4, distribution = 'bernoulli',
                    n.cores=5), error = function(e){e}
      )

    if(!inherits(temp, 'error')){
        brt.best.iter <- gbm.perf(temp, method="OOB")
        brt.preds <- prediction(predict(temp, newdata=test,
                                                    n.trees=brt.best.iter), prezTE)
        brt.perf <- performance(brt.preds,"tpr","fpr")
        brt.auc[i] <- unlist(performance(brt.preds, 'auc')@y.values)
    }
    if(inherits(temp, 'error')){
        brt.auc[i] <- NA
    }

    print(i)
    gc()
   }
    return(data.frame(baseline = baseline.auc, brt = brt.auc))
  }


#sample 50 random parasites from those with 40 or more occurrence points
psites <- table(pest$P_SP)
psites <- psites[psites > 40]

set.seed(30)
psites <- names(psites)[sample(1:length(psites), 50)]


auc5 <- getBackground(psites, 5)
auc10 <- getBackground(psites, 10)
auc20 <- getBackground(psites, 20)
auc40 <- getBackground(psites, 40)


aucMns <- c(mean(auc5$brt), mean(auc10$brt), mean(auc20$brt), mean(auc40$brt))
aucSd <-  c(sd(auc5$brt), sd(auc10$brt), sd(auc20$brt), sd(auc40$brt))

plot(x=c(5,10,20,40), aucMns, type='b', ylab='Model accuracy (AUC)', xlab='Proportion background points (x presence points)', ylim=c(0.75,1), lwd=2, tck=0.01, las=1, xaxt='n')
points(x=10, y=aucMns[2], pch=21, bg='dodgerblue', cex=1.1)
axis(1, c(5,10,20,40), tck=0.01 )
lines(c(5,10,20,40),aucMns+aucSd, lty=2, lwd=2)
lines(c(5,10,20,40),aucMns-aucSd, lty=2, lwd=2)

dev.copy(pdf, 'backgroundPts.pdf', height=5, width=5.75); dev.off()

```




### Diagnostic PCA plot (scree plot)

```{r}

parPCA <- princomp(intMat.trunc)
# Eigenvalues
parEig <- (parPCA$sdev)^2
# Variances in percentage
parVarEig <- parEig*100/sum(parEig)
# Cumulative variances
parCumVar <- cumsum(parVarEig)
cumVarPrint <- as.character(substr(parCumVar[1:10], start=1, stop=4))
cumVarPrint[1] <- '9.0'

par(mar=c(4,4,0.5,0.5))
barplot(parVarEig[1:10], las=1,
  col=add.alpha(c(rep('dodgerblue',5), rep('firebrick', 5)), 0.5),
  border=c(rep('dodgerblue',5), rep('firebrick', 5)),
  main='', ylim=c(0,15), xaxt='n', xlab='Principal component axis',
  tck=-0.01, ylab='Percent variance explained')
axis(1, labels=c(1:10), at=seq(0.8,11.5, length.out=10), tck=-0.01)

text(x=seq(0.7,11.5, length.out=10), y = parVarEig[1:10]+0.5,
  labels=cumVarPrint)

dev.copy(pdf, 'scree.pdf', height=3, width=6); dev.off()

```
